A cautionary note on the use of Ornstein Uhlenbeck and other models in macroevolutionary studies

### 2015-04-09 Manuscript/ohyou.tex
Edits to intro and appendix

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{enumerate}     
\pagenumbering{arabic}

@@

  Phylogenetic comparative methods are powerful tools for identifying patterns in the evolution of species traits, and for potentially inferring the evolutionary processes that underlie them \citep[e.g.,][]{freckleton2009seven,Nunn:2011aa,o2012evolutionary,pennell2013integrative}. 
  These approaches have been used, for example, to model niche conservatism \citep{Wiens:2010aa}, infer potential rates of species responses to climate change \citep{Quintero:2013aa}, test the role of ecological niche as a driver of morphological evolution \citep{pienaar2013macroevolution}, compare rates of evolution of species traits \citep{Claramunt:2012ab} and test for constraints in adaptive radiations \citep{blackburn2013adaptive}. % Maybe cut to three examples?
  %Unfortunately, although widely used, the properties of some phylogenetic comparative methods are poorly understood leading to the potential for inappropriate use and misinterpretation of results.

@@

  Here we investigate the OU model in more detail. 
  We present simulations demonstrating the inherent bias in estimating the $\alpha$ parameter, discuss the intricacies of interpreting OU models biologically, and provide advice for appropriate use of OU models in phylogenetic comparative analyses. We also show that very small amounts of measurement error in data can have profound effects on the performance of models. 
  Many of our findings are applicable to other models of evolution (see Discussion and Supplemental Figs. S6-S11) 
  % Add figure refs here
  , but we focus on the OU model because of the recent increase in publications using the model (see below) and because of the under-appreciated ambiguity in the link between pattern and process when interpreting estimates of the $\alpha$ parameter. 
  We are not the first to mention these problems, but we hope by summarising the issues and giving recommendations we can help those less familiar with the technical literature \citep[e.g.,][]{ho2013asymptotic}.

  \subsection{What is the OU model?}
    According to the Brownian model \citep{cavalli1967,felsenstein1973maximum}, a trait X evolves at random at a rate $\sigma$:

@@

    Note that this model has two parameters in addition to those of the Brownian model: $\alpha$ and $\mu$. 
    The parameter $\mu$ is a long-term mean, and it is assumed that species traits evolve around this value. 
    $\alpha$ is the strength of evolutionary force that returns traits back towards the long-term mean if they evolve away from it. 
    $\alpha$ is sometimes referred to as the ``rubber band'' parameter because of the way it pulls traits back towards $\mu$. 
    When $\alpha$ is close to zero, evolution is approximately Brownian, then as $\alpha$ gets larger (around 0.5 in our simulations) the non-Brownian behaviour of the model starts to become apparent. 
    Eventually, when $\alpha$ is really large (i.e., greateer than XXX), all imprint of history is lost and the evolution is essentially a rapid burst at the present (FIGURE 2). 
    This will be indistinguishable from phylogenetically unstructured data. % needs better phrasing
    % Fix the figure or send to supplemental

  \subsection{What do people use the OU model for?}
    The popularity of the OU model has grown exponentially in recent years (Fig. \ref{figure.literature}), in part because these models are now easy to implement apply via packages in R \citep[e.g. ouch, GEIGER and OUwie;][]{Butler:2004aa,Harmon:2008aa,beaulieu2012ouwie}. 
    OU models have become particularly prevalent in ecological studies: 829 papers containing the phrases ``Ornstein Uhlenbeck'' and ``ecology'' were published between 2012 and 2014 (Google Scholar search 15th March 2015; see Supplementary Material). % Need to add ref to supp mat
    Additionally, although they are pattern based, fit to an OU model is now being used as evidence for the action of ecological, or ecologically driven, processes such as phylogenetic niche conservatism, convergent evolution and stabilizing selection \citep[e.g.,][]{Wiens:2010aa,christin2013anatomical,ingram2013surface}. OU models are also increasingly used in palaeontological studies \citep[e.g.,][]{slater2013,soul2015taxonomy}.

@@

\section{Materials and Methods}
  \subsection{Maximum Likelihood simulations}
  \label{section:sims.methods} 
    We used simulations to explore the statistical properties of $\alpha$. 
    We simulated phylogenies with 25, 50, 100, 150, 200, 500, or 1000 tips under pure birth, constant-rate Birth-Death (extinction fractions of 0.25, 0.5 and 0.75), or temporally varying speciation rate (speciation rate modeled as time from the root raised to the power 0.2, 0.5, 2 and 5) models. 
    We simulated 1000 phylogenies for each combination of tips and models resulting in 56,000 simulated phylogenies in total. 
    Trees were simulated using the R package TESS \cite{hohna2013fast}. 
    We then simulated the evolution of a single trait under a Brownian motion model on each phylogeny using the R package MOTMOT \citep{Thomas:2011aa}. 
    We estimated $\alpha$ and compared the fit of a Brownian model to that of an OU model using a likelihood ratio test with 1 degree of freedom with the \texttt{transformPhylo.ML} function in MOTMOT \citep{Thomas:2011aa}.

@@

    Figure \ref{figure.profiles} shows some examples of profile likelihoods for α generated on different datasets that were simulated according to a Brownian model. It is clear that the shape of the tree affects the shape of the likelihood profile. Trees with increasing speciation rates from root to tip (i.e., that have a lot of recently evolved species) generate very wide, flat likelihood profiles (Fig. \ref{figure.profiles}A). On the other hand, trees with declining speciation rates from root to tip (Fig. \ref{figure.profiles}B), or generated according to a Yule process (Fig. \ref{figure.profiles}C), yield more strongly peaked profiles.

@@

The effects of the asymmetric profile likelihoods become clear when calculating Type I error for the OU model based on ML estimates across a range of tree shapes and sizes (Fig. \ref{figure.rejection}; see Supplemental Figs. S2 and S3 for results with alternative priors for the Bayesian analyses). Regardless of tree shape, Type I error rates are unacceptably high when tree size is small (Figs. \ref{figure.rejection}A, \ref{figure.rejection}C and \ref{figure.rejection}E). For some tree shapes (particularly where speciation rates accelerate towards the present), Type I error remains $>0.05$ even for trees with 1000 tips. Bayesian estimation is not subject to bias in the log-likelihood statistic and consistently rejects the OU model. However, estimates of α are similar regardless of method.

@@

The biases reported above are not unique to the OU model. The parameter $\delta$, which tests for accelerating or decelerating rates of evolution \citep{Pagel:1997aa,Pagel:1999aa} also suffers from elevated Type I error and is upwardly biased, favoring models that imply temporally increasing evolutionary rates (Supplemental Fig. S8). In contrast, although the speciational model ($\kappa$) has slightly elevated Type I errors, the parameter estimates are unbiased for the majority of tree sizes and shapes (Supplemental Fig. S6). As previously shown (Freckleton et al. 2002), the $\lambda$ model is conservative with low Type I error rates (Supplemental Fig. S7). As with the OU model, the $\delta$ models are all strongly favored over Brownian motion when there is error in the species data (Supplemental Figs. S9-S11).

@@

Interpreting results
 - OU is favoured over BM. If alpha is small it's basically BM. What
is alpha? Talk about half lives.
 - Not stabilising selection
 - Not an issue if you're just mopping up error structure for PGLS (?)
 - If really want to be certain OU over BM you need 200 species
 - Account for measurement error

@@

\section{Appendix}
\subsection{Model Outline} % correct the equations!
\label{section:models} 
According to the Brownian model, a trait X evolves at random at a rate $\sigma$:

  \begin{equation}
    dX(t) = \sigma dW(t)
    %\label{equation:BMrate} 
  \end{equation}

@@

  \begin{equation}
    dX(t) = - \alpha (X(t) - \mu) + \sigma dW(t)
    %\label{equation:OUrate} 
  \end{equation}

The parameter $\mu$ is a long-term mean, and it is assumed that species evolve around this value. $\sigma$ is the strength of evolutionary force that returns traits back towards the mean if they evolve away. The OU model was introduced to population genetics by \citet{Lande:1976aa} to model stabilizing selection in which the mean was recast as a fitness optimum on an adaptive landscape. The process operating in comparative data is analogous, although clearly is not stabilizing selection despite being sometimes referred to as such (see Discussion). This model has two parameters in addition to those of the Brownian model, $\alpha$ and $\mu$.
The OU model predicts that after a time $T$ for a species $i$, the variance in trait value $X_i$ is:

  \begin{equation}
    var(X_i) = 1 - e^{-2 \alpha T}
    \label{equation:OUvar} 
  \end{equation}

@@

  \begin{equation}
    cov(X_i, X_j) = e^{-2 \alpha (T - t_{ij})} (1 - e^{-2\alpha t_{ij}})
    \label{equation:OUcov} 
  \end{equation}

The variances and covariances predicted by equations \ref{equation:OUvar} and \ref{equation:OUcov} are more complex than those predicted by the Brownian model. In the light of the results below, some properties of this model are worth highlighting:

\begin{enumerate}[(i)]

  \item If $\alpha$ is small then evolution is approximately Brownian: If $\alpha$ is small then $1 - e^{-2\alpha T} \approx 2\alpha T$, i.e. traits accrue variance as if evolving according to a Brownian process. The question is obviously what is a small value of $\alpha$? Figure \ref{figure.traitsim} provides some guidance on this issue, and see point (iv) below.\\

  \item If species $i$ and $j$ diverged recently, evolution is approximately Brownian: if two species diverged recently, then $T - t_{ij} \approx 0$ and hence $cov(X_i, X_j) \approx 1 - e^{-2\alpha t_{ij}} \approx 2\alpha t_{ij}$. Thus, recently diverged species provide little information relevant to estimating non-Brownian evolution according to an OU process. \\

  \item In the long-term the imprint of history is weakened: if $T$ is large (i.e. evolution proceeds for a long time), equation ??? predicts that the variance in $X_i$ tends to a constant, i.e. because the expected value of $X_i$ is $\mu$. Similarly, in equation (4), the covariance between traits tends to a constant because $T$ becomes large relative to $t_{ij}$. Consequently for large groups the model implies that the imprint of history is weak. As a result adding more species to a phylogeny will contribute only relatively weakly to fitting the model, as the deeper branches within a phylogeny will contribute little information to estimating the parameters.\\

\end{enumerate}

Figure \ref{figure.traitsim} gives some numerical examples, chosen to outline the behavior of $\alpha$ in relation to the above. We show the predicted covariance in traits \ref{equation:OUcov} scaled by the overall variance \ref{equation:OUvar} plotted against the time since divergence, $t_{ij}$ scaled by tree height. Figure \ref{figure.traitsim} highlights that the behavior of $\alpha$ can be classified into three regions. When $\alpha$ is low the model behaves very much like Brownian motion, with the covariance scaling approximately linearly with time. Only as the value becomes moderately larger (in this case approaching a value of $0.5$), does the non-Brownian behavior become apparent. Secondly, there is a region of moderate values in which the model leads to markedly non-Brownian patterns of trait variation (Fig. \ref{figure.traitsim}B).  Finally, when $\alpha$ is large the model behaves differently again: in this case, all imprint of history is lost and the evolution is essentially a rapid burst at the present (Fig. \ref{figure.traitsim}C). This will be indistinguishable from phylogenetically unstructured data.

\section{Supplementary Material}

  \subsection{Literature Review}
  \label{section:literature.methods} 
  To get an overview of the use of OU models in ecology, evolution and palaeontology, we used Google Scholar (accessed 13th March 2015) to locate papers published between 2005 \citep[when the R package ouch was released;][]{Butler:2004aa} and 2014 that contained the terms ``Ornstein Uhlenbeck'' and either ``'ecology'', or ``evolution'' and ``biology'' (the ``biology'' term was added to omit physics papers which also use the term ``evolution''), or ``paleo/palaeo''. We also recorded the total number of papers containing the terms ``ecology'', or ``evolution'' and ``biology'', or ``paleo/palaeo'', published between 2005 and 2014 and plotted the number of OU papers published each year as a proportion of the total number of papers published (Fig. \ref{figure.literature}).

  Next we filtered our Google Scholar search results to focus on empirical papers using OU models (rather than methods papers) published in the following journals: The American Naturalist, Ecology Letters, Evolution, Journal of Evolutionary Biology, Nature, Proceedings of the National Academy of Sciences, USA, Proceedings of the Royal Society B: Biological Sciences, and Science. For each of these papers we recorded the number of species in the analysis, the study group (amphibians, birds, fish, mammals, reptiles, invertebrates or plants), the statistical package or specific R package used to fit the models, and the reason the authors state for using an OU model (ancestral state reconstructions, detecting convergent evolution, controlling for phylogeny, selecting a model of trait evolution, or other). Where papers included multiple analyses using different numbers of species we used the median number of taxa. Where papers had multiple study groups, statistics/R packages or reasons for fitting OU models we counted them in each relevant category. We summarise these results in Figure S1 and the full dataset is available in Supplemental Table S1 along with the full list of references.

  \label{section:literature.results}
  In total, 3720 papers containing the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” were published between 2005 and 2013, and the number has increased substantially since 2005 (Fig. \ref{figure:literature}). Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the “best” model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

>>>>

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{enumerate} 
%\usepackage[nomarkers, nolists]{endfloat}  % To push floats to the end   
\pagenumbering{arabic}

@@

  Phylogenetic comparative methods are powerful tools for identifying patterns in the evolution of species traits, and for potentially inferring the evolutionary processes that underlie them \citep[e.g.,][]{freckleton2009seven,Nunn:2011aa,o2012evolutionary,pennell2013integrative}. 
  These approaches have been used, for example, to infer potential rates of species responses to climate change \citep{Quintero:2013aa}, test the role of ecological niche as a driver of morphological evolution \citep{pienaar2013macroevolution}, and test for constraints in adaptive radiations \citep{blackburn2013adaptive}. 
  %Unfortunately, although widely used, the properties of some phylogenetic comparative methods are poorly understood leading to the potential for inappropriate use and misinterpretation of results.

@@

  Here we investigate the OU model in more detail. 
  We present simulations demonstrating the inherent bias in estimating the $\alpha$ parameter, discuss the intricacies of interpreting OU models biologically, and provide advice for appropriate use of OU models in phylogenetic comparative analyses. We also show that very small amounts of measurement error in data can have profound effects on the performance of models. 
  Many of our findings are applicable to other models of evolution (see Discussion and Supplemental Figs. S6-S11) 
  % Add figure refs here
  , but we focus on the OU model because of the recent increase in publications using the model and because of the under-appreciated ambiguity in the link between pattern and process when interpreting estimates of the $\alpha$ parameter. 
  We are not the first to mention these problems, but we hope by summarising the issues and giving recommendations we can help those less familiar with the technical literature \citep[e.g.,][]{ho2013asymptotic,ho2014intrinsic,boettiger2012your}. %ives garland 2010, hansen bartoszek et al

\section{Materials and Methods}
  \subsection{What is the OU model?}
    According to the Brownian model \citep{cavalli1967,felsenstein1973maximum}, a trait X evolves at random at a rate $\sigma$:

      \begin{equation}
        dX(t) = \sigma dW(t)
      \end{equation}

@@

      \begin{equation}
        dX(t) = - \alpha (X(t) - \mu) + \sigma dW(t)
      \end{equation}

    Note that this model has two parameters in addition to those of the Brownian model: $\alpha$ and $\mu$. 
    The parameter $\mu$ is a long-term mean, and it is assumed that species traits evolve around this value. 
    $\alpha$ is the strength of evolutionary force that returns traits back towards the long-term mean if they evolve away from it. 
    $\alpha$ is sometimes referred to as the ``rubber band'' parameter because of the way it pulls traits back towards $\mu$. For more details see Appendix.

    When $\alpha$ is close to zero, evolution is approximately Brownian, then as $\alpha$ gets larger the non-Brownian behaviour of the model starts to become apparent. 
    Eventually, when $\alpha$ is really large, all imprint of history is lost and the trait evolution is essentially a rapid burst at the present (FIGURE 2).
    We discuss how to interpret values of $\alpha$ below.

    % Fix the figure or send to supplemental

  \subsection{What do people use the OU model for?}
    The popularity of the OU model has grown extensively in recent years (Fig. \ref{figure.literature}), in part because these models are now easy to implement apply via packages in R \citep[e.g. ouch, GEIGER and OUwie;][]{Butler:2004aa,Harmon:2008aa,beaulieu2012ouwie}.

    OU models have become particularly prevalent in ecological studies: 829 papers containing the phrases ``Ornstein Uhlenbeck'' and ``ecology'' were published between 2012 and 2014 (Google Scholar search 15th March 2015; see Supplementary Material). % Need to add ref to supp mat
    Additionally, although they are pattern based, fit to an OU model is now being used as evidence for the action of ecological, or ecologically driven, processes such as phylogenetic niche conservatism, convergent evolution and stabilizing selection \citep[e.g.,][]{Wiens:2010aa,christin2013anatomical,ingram2013surface}. OU models are also increasingly used in palaeontological studies \citep[e.g.,][]{slater2013,soul2015taxonomy}.

@@

     % OK somewhere here need to clarify that a) most ou studies are looking at fitting the model, and comparing models, not using multiple optima stuff. Then clarify that here we are focussing on that popular use of the

     % Problems with the OU model and where they are reported in the literature?

     % This maybe belongs in the main text in shorter form  
  Figure \ref{figure.traitsim} gives some numerical examples, chosen to outline the behavior of $\alpha$ in relation to the above. We show the predicted covariance in traits \ref{equation:OUcov} scaled by the overall variance \ref{equation:OUvar} plotted against the time since divergence, $t_{ij}$ scaled by tree height. Figure \ref{figure.traitsim} highlights that the behavior of $\alpha$ can be classified into three regions. When $\alpha$ is low the model behaves very much like Brownian motion, with the covariance scaling approximately linearly with time. Only as the value becomes moderately larger (in this case approaching a value of $0.5$), does the non-Brownian behavior become apparent. Secondly, there is a region of moderate values in which the model leads to markedly non-Brownian patterns of trait variation (Fig. \ref{figure.traitsim}B).  Finally, when $\alpha$ is large the model behaves differently again: in this case, all imprint of history is lost and the evolution is essentially a rapid burst at the present (Fig. \ref{figure.traitsim}C). This will be indistinguishable from phylogenetically unstructured data.

  \subsection{Maximum Likelihood simulations}
  \label{section:sims.methods} 
    We used simulations to explore the statistical properties of $\alpha$. 
    We simulated phylogenies with 25, 50, 100, 150, 200, 500, or 1000 tips under pure birth, constant-rate Birth-Death (extinction fractions of 0.25, 0.5 and 0.75), or temporally varying speciation rate (speciation rate modeled as time from the root raised to the power 0.2, 0.5, 2 and 5) models. 
    We simulated 1000 phylogenies for each combination of tips and models resulting in 56,000 simulated phylogenies in total. 
    Trees were simulated using the R package TESS \cite{hohna2013fast}. 
    We then simulated the evolution of a single trait under a Brownian motion model on each phylogeny using the R package MOTMOT \citep{Thomas:2011aa}. 
    We estimated $\alpha$ and compared the fit of a Brownian model to that of an OU model using a likelihood ratio test with 1 degree of freedom with the \texttt{transformPhylo.ML} function in MOTMOT \citep{Thomas:2011aa}.

@@

    Figure \ref{figure.profiles} shows some examples of profile likelihoods for $\alpha$ generated on different datasets that were simulated according to a Brownian model. It is clear that the shape of the tree affects the shape of the likelihood profile. Trees with increasing speciation rates from root to tip (i.e., that have a lot of recently evolved species) generate very wide, flat likelihood profiles (Fig. \ref{figure.profiles}A). On the other hand, trees with declining speciation rates from root to tip (Fig. \ref{figure.profiles}B), or generated according to a Yule process (Fig. \ref{figure.profiles}C), yield more strongly peaked profiles.

@@

The effects of the asymmetric profile likelihoods become clear when calculating Type I error for the OU model based on ML estimates across a range of tree shapes and sizes (Fig. \ref{figure.rejection}; see Supplemental Figs. S2 and S3 for results with alternative priors for the Bayesian analyses). Regardless of tree shape, Type I error rates are unacceptably high when tree size is small (Figs. \ref{figure.rejection}A, \ref{figure.rejection}C and \ref{figure.rejection}E). For some tree shapes (particularly where speciation rates accelerate towards the present), Type I error remains $>0.05$ even for trees with 1000 tips. Bayesian estimation is not subject to bias % check in the log-likelihood statistic and consistently rejects the OU model. However, estimates of α are similar regardless of method.

@@

The bias % check
 reported above are not unique to the OU model. The parameter $\delta$, which tests for accelerating or decelerating rates of evolution \citep{Pagel:1997aa,Pagel:1999aa} also suffers from elevated Type I error and is upwardly biased, favoring models that imply temporally increasing evolutionary rates (Supplemental Fig. S8). In contrast, although the speciational model ($\kappa$) has slightly elevated Type I errors, the parameter estimates are unbiased for the majority of tree sizes and shapes (Supplemental Fig. S6). As previously shown (Freckleton et al. 2002), the $\lambda$ model is conservative with low Type I error rates (Supplemental Fig. S7). As with the OU model, the $\delta$ models are all strongly favored over Brownian motion when there is error in the species data (Supplemental Figs. S9-S11).

@@

Interpreting results
 - OU is favoured over BM. If alpha is small it's basically BM. What
is alpha? Talk about half lives. add in ives and garland advice on what is big and what is small here?
 - Not stabilising selection
 - Not an issue if you're just mopping up error structure for PGLS (?)
 - If really want to be certain OU over BM you need 200 species
 - Account for measurement error

@@

\section{Appendix}
  \label{section:models}
  \setcounter{equation}{0}

  According to the Brownian model, a trait X evolves at random at a rate $\sigma$:

@@

  The parameter $\mu$ is a long-term mean, and it is assumed that species evolve around this value. 
  $\sigma$ is the strength of evolutionary force that returns traits back towards the mean if they evolve away. 
  This model has two parameters in addition to those of the Brownian model, $\alpha$ and $\mu$.
  The OU model predicts that after a time $T$ for a species $i$, the variance in trait value $X_i$ is:

% Still not sure of these two equations!
    \begin{equation}
      var(X_i) = \frac{(1 - e^{-2 \alpha T}) \times \sigma^2}{2 \alpha}
      \label{equation:OUvar} 
    \end{equation}

@@

    \begin{equation}
      cov(X_i, X_j) = \frac{e^{-2 \alpha (T - t_{ij})} \times (1 - e^{-2\alpha t_{ij}})\times \sigma^2}{2 \alpha}
      \label{equation:OUcov} 
    \end{equation}

  The variances and covariances predicted by equations \ref{equation:OUvar} and \ref{equation:OUcov} are more complex than those predicted by the Brownian model. 
  In the light of the results above, some properties of this model are worth highlighting:

  \begin{enumerate}[(i)]
    \item If $\alpha$ is small then evolution is approximately Brownian: If $\alpha$ is small then $1 - e^{-2\alpha T} \approx 2\alpha T$, i.e. traits accrue variance as if evolving according to a Brownian process.\\

    \item If species $i$ and $j$ diverged recently, evolution is approximately Brownian: if two species diverged recently, then $T - t_{ij} \approx 0$ and hence $cov(X_i, X_j) \approx 1 - e^{-2\alpha t_{ij}} \approx 2\alpha t_{ij}$. 
    Thus, recently diverged species provide little information relevant to estimating non-Brownian evolution according to an OU process.\\

    \item In the long-term, the imprint of history is weakened: if $T$ is large (i.e. evolution proceeds for a long time), equation \ref{equation:OUvar} predicts that the variance in $X_i$ tends to a constant, i.e. because the expected value of $X_i$ is $\mu$. 
    Similarly, in equation \ref{equation:OUrate}, the covariance between traits tends to a constant because $T$ becomes large relative to $t_{ij}$.
    Consequently for large groups the model implies that the imprint of history is weak.\\
  \end{enumerate}

\input{ohyou_suppmat}
